A pipeline to analyse time-course gene expression data

### 2020-07-30 scripts/manuscript.Rmd
Edits + rebase here and there

keywords: >
    time-course gene expression data, clustering, differential expression,
    workflow
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
bibliography: bibliography.bib
---

```{r echo=FALSE, results="hide", message=FALSE}
# Loading dependencies
library(devtools)
library(limma)
library(splines)
library(stats)
library(knitr)
library(BiocStyle)
library(biomaRt)
library(kableExtra)

@@

library(topGO)

@@

The manuscript was last compile on `r Sys.time()`.

@@

 `moanin` is available from GitHub, and can be installed using the `install_github` function in the package `devtools`, along with the corresponding package that contains time course datasets we will use:

```{r eval=FALSE}
library(devtools)
install_github("NelleV/moanin")
install_github("NelleV/timecoursedata")
```

@@

```{r eval=FALSE}
BiocManager::install(c("NMF","ggfortify","topGO","biomaRt","KEGGprofile","BiocWorkflowTools"))
```

@@

This workflow is illustrated using data from a micro-array time-course
experiment, exposing mice to three different strains of influenza, and
collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later)
[@shoemaker:ultrasensitive]. The three strains of influenza used in the
study are (1) a low pathogenicity seasonal H1N1 influenza virus
(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
$10^5$ PFU of each virus. An adidtional 42 mice were injected with a lower dose
of the Vietnam avian influenza virus ($10^3$ PFU).

By combining gene expression time-course data with virus growth data, the
authors show that the inflammatory response of lung tissue is gated until a
threshold of the virus concentration is  exceeded in the lung. Once this
threshold is exceeded, a strong inflammatory and cytokin production occurs.
This results provides evidence that the pathology response is non-linearly
regulated by virus concentration.

@@

```{r echo=FALSE}
kable(head(meta))
```

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots. We also reorder the factor `Group` which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).

@@

Before performing any additional exploratory analysis, let us only keep highly
variable genes: for this step, we keep only the top 50% most variable genes.

@@

We can already see interesting patterns emerging from the correlation plot.
First, the cross-correlation amongst samples taking from the control mice is
higher than the cross correlation amongst the rest of the treatments. Second,
the influenza-infected mice mildly react until time point 36. Third, the less
pathogenic the strain is, the closer the samples are to the control condition.
Fourth, the Vietnam samples at time point 120 and 168 are the one that are the
most different from control samples.

@@

In both
situations, we first need to set up a object (a `Moanin` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, ...). The `Moanin` object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the `Moanin` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatement effect, and a
second one named `Timepoint` containing the timepoint information.

```{r echo=FALSE}
kable(head(meta))
```

@@

```{r echo=FALSE}
kable(weekly_de_analysis[1:10, 1:3 ], "latex", digits=3, booktabs=TRUE)

@@

However, the distribution of the number of genes found differentially
expressed by considering each time-point independently highlights the
challenges of such approach. We can see that some timepoints have many less
genes found significantly differentially expressed (e.g. timepoint 6H and 18H
of the Kawasaki strain). While there may be biological differences at those
time points for some genes, it seems unlikely that the large majority of genes
differentially expressed at timepoint 3H stop being differentially expressed
at 6H and then jump back to being differentially expressed at 9H. A more
likely explanation is that there is are some technical or biological artifact
about the samples for 6H that is creating higher variation and thus less
ability to detect significance.

@@

The number of genes found differentially expressed ranges from around 12000 to
29000 depending on the strain and dosage of influenza virus given to the mice.
This corresponds to between 30% to 70% of the genes found differentially
expressed in this time-course experiment.

@@

```{r echo=FALSE}
kable(
    head(log_fold_change_timepoints[, colnames(log_fold_change_timepoints)[1:5]]),
    format="latex",
    booktabs=TRUE)
```

@@

Note, however, that when a gene is not consistently up or down-regulated the estimation
of the direction will not represent accurately the changes observed.

@@

```{r display_lfc, echo=FALSE}
kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE)
```

In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary of each of these methods against each other. We see that each methods captures different elements of
the time-course data, for example, overall change versus the largest change.

@@

With a single measure of log-fold change and the p-value, we can now look
at the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the "timecourse" method of calculating log fold change.

@@

The very large number of genes found differentially expressed impair any
interpretation one would attempt: with 70\% of the genome found differentially
expressed, all pathways are affected by the treatment. Hence the next step of
the workflow to cluster gene expression according to their dynamical response
to the treatment.

@@

To do this, we first
aggregate all p-values obtained during the time-course differential expresison
step in a single p-value using Fisher's method [@fisher:statistical] (`pvalues_fisher_method`).
Next we
select all the genes which have a Fisher-adjusted p-value below 0.05 and a log-fold 
change of at least two for at least one condition and one time-point.

@@

After filtering, we are left with `r dim(de_moanin_model)[1]` genes. We can then apply a
clustering routine. As observed by looking at genes found differentially expressed, 
many genes share a similar gene expression pattern, but on different scale.

@@

We then use the `plot_splines_data`  function, only now applied to the centroids, to visualize the centroids of
each cluster obtained with the splines k-means model.

```{r fig.height=6, fig.width=6, fig.cap="K-means centroids"}
plot_splines_data(de_moanin_model,
    data=kmeans_clusters$centroids, 
    colors=ann_colors$Group,
    smooth=TRUE)
```

These centroids are on a 0-1 scale, because of our rescaling of the spline
fits, and do not represent the actual gene expression level. In Figure \ref{fig:genesVsCentroids}, we plot a few
of the actual genes assigned to cluster 2 to see the difference:

```{r genesVsCentroids, fig.height=6, fig.width=6, fig.cap="Genes from cluster 2"}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
dataToPlot = rbind(
    assay(de_moanin_model[cluster2Genes[3:6],]), 
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
plot_splines_data(de_moanin_model, data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
    mar=c(1.5,2.5,2,0.1))
```

The spline plots are fitted per gene in the above plot and illustrate that
the genes assigned to a cluster have differing degrees to which their data
fits that of the cluster centroid, and indeed fits that of their individual
spline fits.

We can also overlay the centroid on the data, after rescaling the data to be between 0-1, like the centroids

```{r}
de_moanin_rescale = de_moanin_model
assay(de_moanin_rescale) = rescale_values(data=assay(de_moanin_model))
plot_splines_data(de_moanin_rescale,  
    centroid=kmeans_clusters$centroids[cluster_of_interest,], 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    subset_data=cluster2Genes[3:6],
    mar=c(1.5,2.5,2,0.1))
```
**EAP** I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this. The `rescale_values` function is too sensitive to outliers. Would be better to rescale each centroid to scale of gene. If we had a helper function to do that, could easily add that to plot_splines_data (made issue on moanin), but probably right now should remove this.

@@

Because there is variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster. Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would
like to have a mechanism to assign all genes to a cluster.

@@

**Assigning cluster labels:** We could just assign each gene to a cluster
based on which cluster gave the minimum score. By default,
`splines_kmeans_score_and_label` does not do that, but rather requires a
sufficiently low enough goodness-of-fit score. The criteria for being
"sufficiently low" is based on looking at the distribution of the scores of
all genes on all clusters (i.e. the entire `scores` matrix returned by
`splines_kmeans_score_and_label`). A gene is then assigned to a cluster only
if their best score is above the 50% percentile of that distribution, with the
remaining genes getting `NA` as their assignment (this choice can be changed
by `proportion_genes_to_label`). Note that because the distribution of scores
of genes across *all* clusters are used, this is not equivalent to assigning
50% of genes to a cluster -- it is possible that all genes are assigned to a
cluster.

@@

```{r visualizeScores, fig.height=6, fig.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
# This time without filtering labels
# We can give the previous calculated scores to `previous_scores` to save time
unfiltered_scores<-splines_kmeans_score_and_label(moanin_model,kmeans_clusters,proportion_genes_to_label=1, previous_scores=scores_and_labels$scores)
best_score = rowMin(scores_and_labels$scores)
best_label = unfiltered_scores$labels

@@

Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in%
genes_in_cluster2)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes (Figure \ref{fig:topScoreInCluster2})

```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
# order them by their score in cluster
ord = order(scores_and_labels$scores[genes_in_cluster2,
				     cluster_of_interest])
cluster2Score = genes_in_cluster2[ord]
dataToPlot = rbind(
    assay(moanin_model[cluster2Score[1:4], ]),
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
plot_splines_data(
    moanin_model, data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
    mar=c(1.5, 2.5, 2, 0.1))
```

@@

### Looking at specific clusters in details.

Now, let us look more in detail some specific clusters. Cluster 6 seems
particularly interesting: it captures genes with strong differences between
the different influenza treatments and the control, while the control remains
relatively flat.

@@

```{r heatmapCluster8, fig.width=6, fig.height=8, fig.cap="Heatmap of genes in Cluster 8"}
layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))

@@

Those two heatmaps demonstrate that the clustering method successfully cluster
genes that are on different scales, and yet share the same dynamical response
to the treatments.

@@

- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
  the elbow methods looks at the total with cluster sum of squares 
  as a function of the number of clusters (WCSS). When adding clusters
  doesn't decrease the WCSS by a sufficient
  amount, one can consider stopping. This method thus provides visual aid to
  the user to choose the number of clusters, but often the "elbow" is hard to
  see on real data, where the number of clusters is not clearly defined.
- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
  refers to a method of validation of consistency within clusters, and
  provides visual aid to choose the number of clusters.
- **Stability methods** Stability methods are more computationally intensive
  than any other method, as they rely on assessing the stability of the
  clustering for every $k$ to a small randomization of the data. The user is
  then invited to choose the number of cluster based on a number of similarity
  measures.

First, let us run the clustering for all values of $k$ of interest. We
will, for each clustering, conserve (1) with within cluster sum of
squares; (2) the clustering assignement (or label) for each gene.

@@

The Elbow method to choose the number of clusters relies on visualization aid
to choose the number of clusters. The method relies on plotting the within
cluster sum of squares (WCSS) as a function of the number of clusters. At some
point, the WCSS will start decreasing more slowly, giving an angle or "elbow"
in the graph.  The number of cluster is chosen at this "Elbow point."

@@

Methods that rely on the stability of the clustering results to choose $k$
thus ensure that the biological interpretation of the clusters hold with
perturbtation to the data. In addition, simulation where the data is generated
with a well defined $k$ show that the clustering is more stable for the
correct of number of the clusters.

Most methods method to find the number of clusters with stability measures
only provide visual aids to guide the user. The first element often visualized
is the consensus matrix: the consensus matrix is an $n \times n$ matrix that
stores the proportion of clustering in which two items are clustered together.
A perfect consensus matrix ordered such as each elements that belong to the
same cluster are adjacent to one another which show blocks along the diagonal
close to 1.

@@

Each column correspond to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 20 resampling runs. For example,
for $k=5$:

```{r,echo=FALSE}
kable(head(stability_5))
```

@@

The model explorer algorithm [@ben-hur:stability] proposes to estimate the
number of clusters exploiting the observation that if the number of clusters
is correct, the clustering results are stable to bootstrap resampling, as
described above.  The distribution of similaries between bootstrapped results
for each $k$ can thus be compared for different values of $k$ and guide the
user in the choice of number of clusters.

@@

From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that $k=5$
is more stable than $k=3$ and $k=4$, but not as stable as $k=2$. The model
explorer strategy, in addition to visualizing the diversity of the centroids,
can thus help assessing an adequate number of clusters.

Now, replot the same model explorer, but only for the clustering experiments
$k=6$, $k=7$, $k=8$, $k=9$ and $k=10$ so that we can see more clearly the stability
measures in that range (Figure \ref{fig:plotModelExplorerZoom}).

@@

The  consensus clustering [@monti:consensus] relies on a similar idea but
instead of looking at the cumulative density of similarity measures of
bootstrapped clustering, the authors suggests plotting the cumulative density
of elements of the consensus matrix. We provide a function
`plot_cdf_consensus` in `moanin` to do this

@@

The stability of the clustering based on the consensus matrix can then be
measured via a single number by looking at the area under the curve: the more
stable the clustering, the closer to 0 or 1 will be the entries of consensus
matrix. The consensus clustering strategy thus suggest at looking at either
the AUC as a function of the number of the clusters or the "improvement" in
the AUC as a function of the number of cluster.

@@

The consensus clustering method suggest that the most stable is $k=2$, which
separates over-expressed genes from under-expressed genes. While it is indeed
a very stable clustering, it does not capture the range of gene expression
patterns present in the data. This shows the limitation of such method on real
data, where the number of clusters is not clearly defined.

@@

Let us first tackle the case of pathway enrichment analysis. We will leverage
the packages `biomaRt` [@durinck:biomart] and `KEGGprofile`
[@zhao:keggprofile] for this step. `KEGGprofile` is a package that facilitates
enrichment analysis on a set of genes labeled with the Ensembl annotation
[@ensembl2020] based on the set of biological pathways described in the KEGG
database[@KEGG].

@@

# Select the columns to print and give nicer column names
pathways = subset(
    pathways,
    select=c(
	"Pathway_Name", "Percentage",
	"pvalueAdj"))
colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
pathways["Percentage"] = pathways["Percentage"] * 100
kable(pathways[1:min(10,nrow(pathways)), ], format="latex", escape=F, digits=4, booktabs=TRUE)
```

@@

kable(go_terms_enriched[1:15, ], format="latex", escape=F,
      digits=3, booktabs=TRUE, row.names=FALSE)
```

@@

This workflow provides a tutoral for the analysis of lengthy time-course gene
expression data in R using the package `moanin`, which aids in implementing common timecourse analyses. We illustrate the workflow through the analysis of mice lung tissue
exposed to different influenza strain and measured over time. 
The proposed workflow covers consists of three common analysis main steps generally performed after quality
control and normalization: (1) differential expression analysis; (2)
clustering of time-course gene expression data; (3) downstream analysis of
clusters. We demonstrate how the use of the package `moanin` allows for easy implementation of these procedures in the setting of time-course data.

@@

All packages use in the workflow are available on GitHub, CRAN, or
Bioconductor.

Data used in this workflow are available from NCBI GEO, accession
[GSE63786](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786).
Normalized data can be found in `moanin`.  Normalization information is
provided as supplementary information.

>>>>

keywords: >
    time-course gene expression data, clustering, differential expression,
    workflow
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
header-includes:
   - \usepackage{float}
output:
    BiocWorkflowTools::f1000_article: default
    ms_document: default
bibliography: bibliography.bib
always_allow_html: yes
---

```{r echo=FALSE, results="hide", message=FALSE}
# Loading dependencies
library(devtools)
library(limma)
library(splines)
library(stats)
library(knitr)
library(BiocStyle)
library(biomaRt)
library(pander)

@@

library(topGO)
library(moanin)
if(!packageVersion("moanin")>="0.99.0-902") stop("Must have version 0.99.0-902 to compile the manuscript")
```

@@

@@

`moanin` and `timecoursedata` are available from bioconductor, and can be
installed using the `install` function in the package `BiocManager`, along
with the corresponding package that contains time course datasets we will use:

```{r eval=FALSE}
library(BiocManager)
# Need to use the development version for now.
BiocManager::install(version="devel")
BiocManager::valid()

BiocManager::install("timecoursedata")
BiocManager::install("moanin")
```

@@

```{r eval=FALSE}
BiocManager::install(
    c("NMF", "ggfortify", "topGO", "biomaRt",
      "KEGGprofile", "BiocWorkflowTools", "timecoursedata", ))
```

@@

This workflow is illustrated using data from a micro-array time-course
experiment, exposing mice to three different strains of influenza, and
collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later)
[@shoemaker:ultrasensitive]. The three strains of influenza used in the
study are (1) a low pathogenicity seasonal H1N1 influenza virus
(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
$10^5$ PFU of each virus. An additional 42 mice were injected with a lower dose
of the Vietnam avian influenza virus ($10^3$ PFU).

By combining gene expression time-course data with virus growth data, the
authors show that the inflammatory response of lung tissue is gated until a
threshold of the virus concentration is  exceeded in the lung. Once this
threshold is exceeded, a strong inflammatory and cytokine production occurs.
This results provides evidence that the pathology response is non-linearly
regulated by virus concentration.

@@

```{r echo=FALSE}
pander(head(meta))
```

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots. We also reorder the factor `Group` which describes
the treatments so that the treatments are ordered from low to high pathogeny
(with Control being first).

@@

Before performing any additional exploratory analysis, let us only keep highly
variable genes: for this step, we keep only the top 50% most variable genes.
In Figure \ref{fig:geneVarHist}, we plot the distribution of variance across
all genes.

@@

We can already see interesting patterns emerging from the correlation plot.
First, the cross-correlation amongst samples taken from the control mice is
higher than the cross correlation amongst the rest of the treatments. Second,
the influenza-infected mice mildly react until time point 36. Third, the less
pathogenic the strain is, the closer the samples are to the control condition.
Fourth, the Vietnam samples at time point 120 and 168 are the one that are the
most different from control samples.

@@

In both situations, we first need to set up an object (a `Moanin` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, ...). The `Moanin`
object will contain information used throughout this analysis, in particular
the condition and timepoints of each sample and the basis matrix for
fitting splines models.

We start by creating the `Moanin` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatment effect, and a
second one named `Timepoint` containing the timepoint information.

```{r echo=FALSE}
pander(head(meta))
```

@@

```{r echo=FALSE}
pander(weekly_de_analysis[1:10, 1:3 ])

@@

However, the distribution of the number of genes found differentially
expressed by considering each time-point independently highlights the
challenges of such approach. We can see that some timepoints have many less
genes found significantly differentially expressed (e.g. timepoint 6H and 18H
of the Kawasaki strain). While there may be biological differences at those
time points for some genes, it seems unlikely that the large majority of genes
differentially expressed at timepoint 3H stop being differentially expressed
at 6H and then jump back to being differentially expressed at 9H. A more
likely explanation is that there are some technical or biological artifacts
about the samples for 6H that are creating higher variation and thus less
ability to detect significance.

@@

The number of genes found differentially expressed ranges from around 12000 to
29000 depending on the strain and dosage of influenza virus given to the mice
(Figure \ref{deTimeDist}).
This corresponds to between 30% to 70% of the genes found differentially
expressed in this time-course experiment.

@@

```{r echo=FALSE}
pander(
    head(log_fold_change_timepoints[,
    colnames(log_fold_change_timepoints)[1:5]]))
```

@@

Note, however, that when a gene is not consistently up- or down-regulated the
estimation of the direction will not accurately represent the changes
observed.

@@

```{r display_lfc, echo=FALSE}
pander(log_fold_change_timecourse[1:5, ])
```

In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary
of each of these methods against each other. We see that each method captures
different elements of the time-course data, for example, overall change versus
the largest change.

@@

With a single measure of log-fold change and the p-value, we can now look at
the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example
of a volcano plot for the comparison of the control to the Kawasaki strain,
using the "timecourse" method of calculating log fold change.

@@

The very large number of genes found differentially expressed impairs any
interpretation one would attempt: with 70\% of the genome found differentially
expressed, all pathways are affected by the treatment. Hence the next step of
the workflow to cluster gene expression according to their dynamical response
to the treatment.

@@

To do this, we first
aggregate all p-values obtained during the time-course differential expression
step in a single p-value using Fisher's method [@fisher:statistical] (`pvalues_fisher_method`).
Next we
select all the genes which have a Fisher-adjusted p-value below 0.05 and a log-fold 
change of at least two for at least one condition and one time-point.

@@

After filtering, we are left with `r dim(de_moanin_model)[1]` genes. We can then apply a
clustering routine. As observed by looking at genes found differentially expressed, 
many genes share a similar gene expression pattern, but on different scales.

@@

We then use the `plot_splines_data`  function, only now applied to the
centroids, to visualize the centroids of each cluster obtained with the
splines k-means model (Figure \ref{fig:centroids}).

```{r centroids, fig.height=6, fig.width=6, fig.cap="K-means centroids"}
plot_splines_data(de_moanin_model,
    data=kmeans_clusters$centroids, 
    colors=ann_colors$Group,
    smooth=TRUE)
```

These centroids are on a 0-1 scale, because of our rescaling of the spline
fits, and do not represent the actual gene expression level. In Figure \ref{fig:genesVsCentroids}, we plot a few
of the actual genes assigned to cluster 2 with the estimated centroid overlaid:

```{r genesVsCentroids, fig.height=6, fig.width=6, fig.cap="Genes from cluster 2"}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
plot_splines_data(de_moanin_model,  
    centroid=kmeans_clusters$centroids[cluster_of_interest,], 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    subset_data=cluster2Genes[3:6],
    mar=c(1.5,2.5,2,0.1))
```

@@

As we can see, while these genes have some similarity with the pattern of the
cluster centroids, these particular genes are not the best examples of the
cluster, in the sense of matching the centroid estimates. Because of this
variability in how well the genes fit a cluster, we would like to be able to
score how well a gene fits a cluster.

Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would
like to have a mechanism to assign all genes to a cluster.

@@

**Assigning cluster labels:** We could just assign each gene to a cluster
based on which cluster gave the minimum score. By default,
`splines_kmeans_score_and_label` does not do that, but rather requires a
sufficiently low enough goodness-of-fit score. The criteria for being
"sufficiently low" is based on looking at the distribution of the scores of
all genes on all clusters (i.e. the entire `scores` matrix returned by
`splines_kmeans_score_and_label`). A gene is then assigned to a cluster only
if their best score is above the 50% percentile of that distribution, with the
remaining genes getting `NA` as their assignment (this choice can be changed
by `proportion_genes_to_label`). Note that because the distribution of scores
of genes across *all* clusters is used, this is not equivalent to assigning
50% of genes to a cluster -- it is possible that all genes are assigned to a
cluster.

@@

```{r visualizeScores, fig.height=6, fig.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
# This time without filtering labels
# We can give the previous calculated scores to `previous_scores` to save time
unfiltered_scores = splines_kmeans_score_and_label(
    moanin_model, kmeans_clusters,
    proportion_genes_to_label=1,
    previous_scores=scores_and_labels$scores)
best_score = rowMin(scores_and_labels$scores)
best_label = unfiltered_scores$labels

@@

Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in%
genes_in_cluster2)` of them remained assigned to cluster `r cluster_of_interest` after our scoring. We can again look at a
few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes (Figure \ref{fig:topScoreInCluster2})

```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
# order them by their score in cluster
ord = order(scores_and_labels$scores[genes_in_cluster2,
				     cluster_of_interest])
cluster2Score = genes_in_cluster2[ord]
plot_splines_data(
    moanin_model, subset_data=cluster2Score[1:4], 
    centroid=kmeans_clusters$centroids[cluster_of_interest,], 
    colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
    mar=c(1.5, 2.5, 2, 0.1))
```
We see that, as expected, these genes are a much better match to the cluster centroids.

@@

### Looking at specific clusters in detail.

Now, let us look more in detail at some specific clusters. Cluster 6 seems
particularly interesting: it captures genes with strong differences between
the different influenza treatments and the control, while the control remains
relatively flat.

@@

```{r heatmapCluster8, fig.width=6, fig.height=8, fig.cap="Heatmap of genes in Cluster 6"}
layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))

@@

Those two heatmaps demonstrate that the clustering method successfully
clusters genes that are on different scales, and yet share the same dynamical
response to the treatments.

@@

- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
  the elbow method looks at the total within cluster sum of squares 
  as a function of the number of clusters (WCSS). When adding clusters
  doesn't decrease the WCSS by a sufficient
  amount, one can consider stopping. This method thus provides visual aid to
  the user to choose the number of clusters, but often the "elbow" is hard to
  see on real data, where the number of clusters is not clearly defined.
- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
  refers to a method of validation of consistency within clusters, and
  provides visual aid to choose the number of clusters.
- **Stability methods** Stability methods are more computationally intensive
  than any other method, as they rely on assessing the stability of the
  clustering for every $k$ to a small randomization of the data. The user is
  then invited to choose the number of clusters based on a number of similarity
  measures.

First, let us run the clustering for all values of $k$ of interest. We
will, for each clustering, conserve (1) with within cluster sum of
squares; (2) the clustering assignment (or label) for each gene.

@@

The Elbow method to choose the number of clusters relies on visualization aid
to choose the number of clusters. The method relies on plotting the within
cluster sum of squares (WCSS) as a function of the number of clusters. At some
point, the WCSS will start decreasing more slowly, giving an angle or "elbow"
in the graph.  The number of clusters is chosen at this "Elbow point."

@@

Methods that rely on the stability of the clustering results to choose $k$
thus ensure that the biological interpretation of the clusters hold with
perturbations to the data. In addition, simulation where the data is generated
with a well defined $k$ show that the clustering is more stable for the
correct of number of the clusters.

Most methods to find the number of clusters with stability measures
only provide visual aids to guide the user. The first element often visualized
is the consensus matrix: the consensus matrix is an $n \times n$ matrix that
stores the proportion of clustering in which two items are clustered together.
A perfect consensus matrix ordered such as all elements that belong to the
same cluster are adjacent to one another which show blocks along the diagonal
close to 1.

@@

Each column corresponds to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 25 resampling runs. For example,
for the first 10 bootstrap samples and $k=5$:

```{r,echo=FALSE}
pander(head(stability_5[, 1:10]))
```

@@

The model explorer algorithm [@ben-hur:stability] proposes to estimate the
number of clusters exploiting the observation that if the number of clusters
is correct, the clustering results are stable to bootstrap resampling, as
described above.  The distribution of similarities between bootstrapped results
for each $k$ can thus be compared for different values of $k$ and guide the
user in the choice of number of clusters.

@@

From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that $k=5$
is more stable than $k=3$ and $k=4$, but not as stable as $k=2$. The model
explorer strategy, in addition to visualizing the diversity of the centroids,
can thus help assess an adequate number of clusters.

Now, replot the same model explorer, but only for the clustering experiments
$k=6$, $k=7$, $k=8$, $k=9$, and $k=10$ so that we can see more clearly the stability
measures in that range (Figure \ref{fig:plotModelExplorerZoom}).

@@

The  consensus clustering [@monti:consensus] relies on a similar idea but
instead of looking at the cumulative density of similarity measures of
bootstrapped clustering, the authors suggest plotting the cumulative density
of elements of the consensus matrix. We provide a function
`plot_cdf_consensus` in `moanin` to do this

@@

The stability of the clustering based on the consensus matrix can then be
measured via a single number by looking at the area under the curve: the more
stable the clustering, the closer to 0 or 1 will be the entries of consensus
matrix. The consensus clustering strategy thus suggests at looking at either
the AUC as a function of the number of the clusters or the "improvement" in
the AUC as a function of the number of clusters.

@@

The consensus clustering method suggest that the most stable is $k=2$, which
separates over-expressed genes from under-expressed genes. While it is indeed
a very stable clustering, it does not capture the range of gene expression
patterns present in the data. This shows the limitation of such methods on real
data, where the number of clusters is not clearly defined.

@@

Let us first tackle the case of pathway enrichment analysis. We will leverage
the packages `biomaRt` [@durinck:biomart] and `KEGGprofile`
[@zhao:keggprofile] for this step. `KEGGprofile` is a package that facilitates
enrichment analysis on a set of genes labeled with the Ensembl annotation
[@ensembl2020] based on the set of biological pathways described in the KEGG
database [@KEGG].

@@

# Select the columns to print and give nicer column names
pathways = subset(
    pathways,
    select=c(
	"Pathway_Name", "Percentage",
	"pvalueAdj"))
colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
pathways["Percentage"] = pathways["Percentage"] * 100
pander(pathways[1:min(10,nrow(pathways)), ])
```

@@

pander(go_terms_enriched[1:15, ])
```

@@

This workflow provides a tutorial for the analysis of lengthy time-course gene
expression data in R using the package `moanin`, which aids in implementing
common timecourse analyses. We illustrate the workflow through the analysis of
mice lung tissue exposed to different influenza strains and measured over time.
The proposed workflow consists of three common analysis main steps
generally performed after quality control and normalization: (1) differential
expression analysis; (2) clustering of time-course gene expression data; (3)
downstream analysis of clusters. We demonstrate how the use of the package
`moanin` allows for easy implementation of these procedures in the setting of
time-course data.

@@

All packages used in the workflow are available on GitHub, CRAN, or
Bioconductor.

Data used in this workflow are available from NCBI GEO, accession
[GSE63786](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786).
Normalized data can be found in `timecoursedata`.  Normalization information is
provided as supplementary information.
